\jaxcosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-05-26 paper.tex
update NUTS/SVI/Neutra paragraphs, Table of ESS. Description of DES-Y1 code exemple of forward/inference modelling using numpyro. Update bib

Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process. \texttt{JAX} offers an easy palatalization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive. Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing. We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.

@@

\subsection{Validation against the Core Cosmology Library (CCL)}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
    \caption{Comparison of matter power spectrum between CCL and jax-cosmo. \textbf{Example of validation plot} \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo/blob/master/docs/notebooks/CCL_comparison.ipynb}{\faFileCodeO}}
    \label{fig:halofit_comparison}
\end{figure}

@@

\subsection{Instantaneous Fisher Forecasts}
\label{sec-fisher-forecast}
\begin{minted}{python}
def likelihood(cosmo):
  # Compute mean and covariance of angular Cls, 
  # for specific probes
  mu, cov = gaussian_cl_covariance_and_mean(cosmo, 
                    ell, probes)
  # Return likelihood value
  return gaussian_log_likelihood(data, mu, cov)

@@

\section{Gradient-based fast cosmological inference}
%
In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo. We have used the \textit{Stochastic Variational Inference} (aka SVI), a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS), and finally NUTS after a \textit{Neural Transport} SVI. All  the methods have been implemented using the \numpyro Probabilistic Programming Language (PPL). 
%
\subsection{Stochastic Variational Inference}
\label{sec-SVI}
%
Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps. If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$. The VI aims to get an approximation of this distribution, ie. $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution. To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
\begin{align}
\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
\end{align}
which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
\begin{equation}
q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
\end{equation}
The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
\begin{equation}
\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
\label{eq-loss-svi-1}
\end{equation}
where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.

Now, one is faced with the evaluation of the expectations during gradient descent. The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}). Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie. a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI. Even, if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to  a NUTS sampler which is described in section \ref{sec-Neural-Reparametrisation} as \textit{Neural Transport}.
%
\subsection{Hamiltonian Monte Carlo}

\subsubsection{Vanilla HMC}

% Show Joe's vanilla HMC results against Cobaya
Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
and developed extensively since.  See \citet{betancourt} for a full review; we describe
very basic features here.

@@

Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.  After the forward model creation for the DES-Y1 3x2pts exercise described in \ref{app-DESY1}, we have performed 
\begin{itemize}
    \item first a reparametrisation of the latent variable uniform priors (Tab.~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$;
    \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$. This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in figure \ref{fig_svi_nuts_4}.
\end{itemize}
We have setup a NUTS sampling using 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to 7 by default (ie. 128 steps for each iteration). The job has been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or NVidia V100 (32GB RAM) at the CCIN2P3 Center.

The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}. The confidential level iso-contours are all in agreement with the highly-optimized \texttt{Cobaya} 
Metropolis-Hastings implementation  
%\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
\citep{2019ascl.soft10019T,2021JCAP...05..057T} (see figure \ref{fig_cobaya_NUTS_SVI}). It is worth to point out that the mass matrix structuring is a very effective when NUTS is run stand-alone as it increases the sampling efficiency by about a factor 2.  The elapse time was 20 hours for the default setting (\texttt{max\_tree\_depth=7}). Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective alternative, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours. This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), so the user is highly encouraged to tune this critical parameter.

%  JEC: the 16 chains X 1000 (max_tree_depth=7) is the run done by Francois. 20 heures sur un GPU TitanXp avec 12 GB de ram 
% 5h sur V100 10 chains x 1000 (max_tree_depth=5)

\begin{figure}
\includegraphics[width=\columnwidth]{figures/Cobaya-NUTS-SVI.pdf}
\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red), and a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green).}
\label{fig_cobaya_NUTS_SVI}
\end{figure}
%
\subsubsection{Neural Transport}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
\caption{In the same conditions of Fig.~\ref{fig_svi_nuts_1} but for the five lens galaxy bias which are highly correlated.}
\label{fig_svi_nuts_4}
\end{figure}

Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI. The Adam optimizer has been used with a learning rate set to $10^{-3}$.  Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.

The SVI optimization has been performed with 20,000 steps to ensure a good convergence of the ELBO loss but no particular tuning of the optimizer has been undertaken to decrease the number of steps. It takes about 2 hours on a Nvidia V100 GPU (32 GB RAM). Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (\texttt{vectorized} option) with 1000 samples each. All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}. 
The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.

The results of the evolution of the 2D parameter contours for some cosmological parameters compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The agreement is quite good even with a rather low number of samples (ie. 200) as shown on figure \ref{fig_svi_nuts_1} and even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters (Fig.~\ref{fig_svi_nuts_4}). It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a worse behaviour.  Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency like the default NUTS settings, without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option). 
%
\subsection{Discussion}
%
The key question is what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
\begin{equation}
    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
\end{equation}
with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation. Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more. At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie. about $1\%$). So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.

>>>>

Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process. \texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive. Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing. We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.

@@

%
\subsection{Validation against the Core Cosmology Library (CCL)}
To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library \citep{2019ApJS..242....2C}. As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function \citep{2012ApJ...761..152T, 2003MNRAS.341.1311S}. 
%
\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
    \caption{Comparison of matter power spectrum between CCL and jax-cosmo. \textbf{Example of validation plot} \nblink{CCL_comparison}}
    \label{fig:halofit_comparison}
\end{figure}

@@

\subsection{Instantaneous Fisher Forecasts}
\label{sec-fisher-forecast}
\begin{minted}[fontsize=\footnotesize]{python}
def likelihood(cosmo):
  # Compute mean and covariance of angular Cls, 
  # for specific probes
  mu, cov = gaussian_cl_covariance_and_mean(cosmo, 
                    ell, probes)
  # Return likelihood value
  return gaussian_log_likelihood(data, mu, cov)

@@

\section{Gradient-based fast cosmological inference}
%
In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo. We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI). All  the methods have been implemented using the \numpyro Probabilistic Programming Language (PPL). 
%
%
\subsection{Hamiltonian Monte Carlo}
%
\subsubsection{Vanilla HMC}
%
% Show Joe's vanilla HMC results against Cobaya
Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
and developed extensively since.  See \citet{betancourt} for a full review; we describe
very basic features here.

@@

Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.  After the forward model creation for the DES-Y1 3x2pts exercise described in \ref{app-DESY1}, we have performed 
\begin{itemize}
    \item first a reparametrisation of the latent variable uniform priors (Tab.~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$;
    \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$. This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_svi_nuts_4}.
\end{itemize}
The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie. 128 steps for each iteration). The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or a NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}. The elapse time for these experiments was 20 hours.

The results in terms of relative effective sample sizes (aka "ESS") are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}. One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya}  Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples. The CL contours are also in very good agreement. It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.   Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours. This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter. 
%
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/Cobaya-NUTS-SVI200.png}
\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
\label{fig_cobaya_NUTS_SVI}
\end{figure*}
\subsection{Stochastic Variational Inference}
\label{sec-SVI}
%
Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps. If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$. The VI aims to get an approximation of this distribution, ie. $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution. To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
\begin{align}
\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
\end{align}
which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
\begin{equation}
q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
\end{equation}
The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
\begin{equation}
\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
\label{eq-loss-svi-1}
\end{equation}
where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.

Now, one is faced with the evaluation of the expectations during gradient descent. The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}). Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie. a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI. As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX. Now even if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
%
\subsubsection{Neural Transport}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.

@@

%\begin{figure}
%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
%\caption{
%Constraints (30\%, 68\%, 90\%) on the five lens galaxy bias. In blue the NUTS sampling described in section \ref{sec-NUTS}, and in red a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
%\label{fig_svi_nuts_4}
%\end{figure}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/Cobaya-NUTS-SVI200_bis.png}
\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
\label{fig_cobaya_NUTS_SVI_bis}
\end{figure}

Using the same DES-Y1 simulation as in previous section, we use a MVN distribution as \textit{guide} to approximate the true posterior (\textit{model}) and feed a SVI. The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$. We have stopped the optimization after 20,000 steps to ensure a good convergence of the ELBO loss, but no particular tuning of the optimizer has been undertaken to decrease the number of steps. It takes 2 hours on the NVidia V100 GPU.

Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler, the chains have run 200 warm samples, and then 200 samples ("low" schema), 1000 samples on a single chain, and finally we have run 10 chains  with 1000 samples each. All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}. The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively. Naturally, more samples lead to better contour precision. But what is illustrative is the "low" schema results as shown in the figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}. Notice the good results even for the highly correlated lens bias parameters. It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.

Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to use these schema to ease the model development in its early phase.
%
\subsection{Discussion}
%
The key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
\begin{equation}
    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
\end{equation}
with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation. Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more. At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie. about $1\%$). So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.
